The persuasion of symbolic elements in government public communications, and their moderation by personal relevance

### 2018-07-01 paper3_relevance and symbols_01.Rmd
Add files via upload

---
title: "Does high personal relevance mitigate the persuasion of symbolic elements in government public communications?"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Previous public administration studies have examined whether public communications can affect citizens trust in government organizations and their policies. Recent experimental studies have demonstrated that these communications can enhance trust not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations (e.g. logos, images and celebrities). Still, under which conditions are citizens more/less likely to be emotionally persuaded by these symbols? Building on the social psychology Elaboration Likelihood Model (ELM), I theorize that citizens are less susceptible to symbols when they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. I put this theory to test through a survey experiment, where I examine circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens' trust in the Environment Protection Ministry's policy plans for reducing the air pollution in a specific area (the Haifa-Bay) and reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The experimental results do not support the hypothesis that higher personal relevance attenuates the effect of symbols. They tentatively suggest that symbols are more likely to persuade citizens who have predominant negative attitudes.*

@@

Public administration scholars have explored whether government organizations can enhance citizens' trust in them by employing public communications. Whereas most of the studies on the effect of government public communications on citizens attitudes has focused on the *informative* aspects of these communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @grimmelikhuijsen_2015; @ho_2017], recent experimental studies further pointed to the emotive effect of *symbolic* elements that are entangled in government communications, such as  agencies' names [@teodoro_2018], brand logos [@karens_2016], and figures in public campaigns [@alonbarkat_2017]. These studies demonstrated that these familiar and well-designed symbolic elements can enhance citizens' trust irratioanlly through a mechanism of "evaluative conditioning" - unconscious transfer of affect from the symbols to the public organizations and policies associated with them [@dehouwer_2012].

While the abovementioned studies have substantiated the distinct effect of symbols, they did not theorize, nor they empirically examined the conditions under which are these symbols are likely to be more/less effective in persuading citizens. This research gap has great significance. From the practical point of view of public organizations, this may enable them to design more effective communications which may enhance citizens' collaboration with their policies. Employing a broader perspective of normative democratic theory, the emotive effect of symbols can be perceived as problematic. It stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Morover, as argued by @alonbarkat_2017, symbolic communication may elicit undue trust in poorly performing organizations, and enable public organizations to escape justified public criticism. Hence, by understanding when citizens are more/less likely to be affected by symbols, we may be able to better assess the potential risk for governments' manipulative use of symbolic communications.

To address this literature gap, I draw upon the social psychology Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011]. I hypothesize that citizens are less susceptible to symbols the more they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. Similarly, higher perceived personal relevance are expected to enhance citizens' inclincation to scrutinize the substantive content of communication. Employing a survey experiment, I test this hypothesis in the context of the Israeli Environment Protection Ministry, and its policy plans for reducing the air pollution in a specific area (the Haifa-Bay) and for reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others.

@@

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" 
<br>

# The moderating role of perceived personal relevance on citizens' responses to symbols and information in communications

How may we apply the theoretical framework of ELM to the study of the effects of symbolic elements in government communications with citizens and their interaction with substantive information?

In sum, I expect that:

@@

The theoretical hypotheses are graphically displayed in **Figure 1**.

<br>
**Figure 1: Theoretical model**

@@

@@

>>>>

---
title: "The persuasion of symbolic elements in government public communications, and their moderation by personal relevance"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Recent experimental studies have demonstrated that government public communications can enhance citizens' trust in government organizations and their policies not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations. Yet, under which conditions are citizens more/less likely to be emotionally persuaded by these symbols? Building on the social psychology Elaboration Likelihood Model (ELM), I theorize that citizens are less susceptible to symbols when they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. I put this theory to test through a survey experiment, where I examine Israeli citizens' trust in the Environment Protection Ministry's policy plans for reducing the air pollution in a specific area (the Haifa-Bay) and for reducing domestic waste. I operationalize the perceived personal relevance of these policies both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The results of both methods do not support the hypothesis that higher personal relevance attenuates the effect of symbols. They tentatively suggest that symbols are more likely to persuade citizens who have predominant negative attitudes.*

@@

A growing public administration literature explores whether government public communications can enhance citizens' trust in public organizations and policies. Whereas most of the studies have focused on the *informative* aspects of government communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @grimmelikhuijsen_2015; @ho_2017], recent experimental studies further pointed to the emotive effect of *symbolic* elements entangled in them, such as  brand names [@teodoro_2018], logos [@karens_2016], and figures [@alonbarkat_2017]. The latter studies suggested that that familiar and well-designed symbols can enhance citizens' trust irrationally through a mechanism of "evaluative conditioning" - unconscious transfer of affect from the symbols to the public organizations and policies associated with them [@dehouwer_2012]. Understanding the effect of symbolic elements is important, insofar as they exist in almost any communications between public organizations and citizens, and most prominently in public campaigns, which have been increasingly used by public-sector organizations [@eshuis_2012; @marland_2017].

While previous studies have substantiated the distinct effect of symbols on citizens' trust, they did not theorize, nor they empirically examined the conditions under which are these symbols are likely to be more/less effective. This research gap has great significance. First, from the practical point of view of government organizations, this may enable them to design more effective communications which may enhance citizens' collaboration with their policies. Second, employing a broader perspective of democratic theory, one can argue that the emotive effect of symbols challenges the fundamental normative expectation that citizens' judgments should be based on their critical thinking. Furthermore, the positive symbolic message can compensate for poor performance and create a misalignment between citizens' perceptions and the "real world" [@alonbarkat_2017]. Hence, by identifying the circumstances under which citizens are more/less likely to be affected by symbols, we would better assess the potential risk for governments' manipulative use of public communications.

To address this literature gap, I draw upon the social psychology Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011]. I hypothesize that citizens are less susceptible to symbols the more they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. Similarly, higher perceived personal relevance are expected to enhance citizens' inclination to scrutinize the substantive content of communication. Employing a survey experiment, I test this hypothesis in the context of the Israeli Environment Protection Ministry, and two of its policy plans: reducing the air pollution in a specific area (the Haifa-Bay); and reducing domestic waste. I operationalize the perceived personal relevance of these policies both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a theoretical framework for investigating the effects of "persuasive" communications on people's attitudes [@petty_1986; @petty_2011]. Whilst this model has been applied in various contexts, thus far, the relevance of ELM has not been directly examined in relation to citizens' responses to government communications.

ELM postulates that people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on the message, they are likely to focus on the arguments in the communication which are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route").[^footnote_dual_process_theories] These two routes for persuasion are regarded as two ends of an "elaboration continuum". When people are located between these two edges, as they are much of the time, persuasion will be determined by a mixture of these processes.

In light of the latter conceptualization of ELM, substantive information in government public communications represent message arguments, which can be critically evaluated based on their logical quality (central route). Conversely, symbolic elements can be regarded as potential peripheral cues, which can affect attitudes by evoking emotions and associations attributed to government organizations and their policies (peripheral route). Accordingly, the more citizens are motivated to think about the content of the communication, and the more they are capable of doing so, the more they are likely to form their attitudes based on its substantive information. Conversely, the less motivated and capable they are to elaborate, the more susceptible they are to symbolic elements in the communication.

[^footnote_dual_process_theories]: The conceptual distinction between the "central and peripheral routes for persuasion" largely resembles the distinctions between heuristic and systematic processing [@chaiken_1980], and systems 1 and 2 [@kahneman_2011].

This literature has pointed to a variety of factors that shape people's motivation and ability to think about the message, and accordingly - their placement along the elaboration continuum and inclination to be persuaded either by the arguments or by peripheral cues. Among others, scholars have theorized that the more people perceive the issue as personally relevant, the more motivated they are to invest in thinking about it, since the consequences of being incorrect are greater. That postulation has been empirically tested, and largely confirmed, by numerous ELM based experimental studies in social psychology and marketing [@petty_1979; @petty_1981; @petty_1983; @johnson_1989; @petty_1990; @burnkrant_1989]. These studies have demonstrated that manipulating the perceived personal relevance of the communication affects subjects' inclination to elaborate, and accordingly, it increases the effect of arguments in the communication and decreases the effect of peripheral cues. Later reviews of ELM literature considered the factor of perceived personal relevance as the most important determinant of motivation to process a message [@petty_2002; @petty_2011].[^footnote_elm_factors]

[^footnote_elm_factors]: For a thorough review of motivational and ability factors affecting likelihood to elaborate see @petty_1998.

Connecting the public administration literature on citizens' responses to government public communications with the above insights from social psychology literature, I theorize that the extent to which citizens perceive the communication as personally relevant moderates the effects of symbolic elements on the one hand and of substantive information on the other hand. I hypothesize that:

@@

These theoretical hypotheses are graphically displayed in **Figure 1**.

<br>
**Figure 1: Theoretical model**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/theoretical_model_relevance.png)

@@

I tested the above mentioned hypotheses via a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans.

@@

Empirically, I focused on the case of the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, both at the national and local levels. I examined citizens' responses to communications of EPM regarding two of its current policy plans. The first policy regards reducing the air pollution in the Haifa-Bay area. This policy issue was selected insofar as it offers a fortunate case for examining the moderating role of perceived personal relevance in a natural and realistic setting. The comparison between those who reside in the polluted area and those who don't represents natural variation in the perceived personal relevance of this policy issue. At the same time, because this policy issue is intangible, invisible, ambiguous and complicated, the differences in personal relevance are not strongly confounded with differences in personal knowledge, as it is the case in many other policy issues [@petty_1983]. Accordingly, those who reside in the polluted Haifa-Bay area are more motivated to scrutinize a communication about the air-pollution in their area of residence, yet are not more capable to do so than those who reside elsewhere. As explained below, the sample population of the experiment consists of residents of Haifa-bay area, matched with residents of cities at the center of Israel with similar aggregate socio-economic profile. The second policy plan regards reducing domestic waste and increasing recycling. This policy issue is assumed to be equally relevant for Haifa-bay residents and others. Therefore, comparing between these two policies enables me to confirm that differences between the two groups, if found, are explained by differences in the perceived personal relevance of the policy issue. For the sake of simplicity, I henceforth refer to these two policy plans as "air-pollution" and "recycling" policies.

EPM also makes a good case for this study due to its significant investment in public communications, and its deployment of various symbolic elements. Here, I focused on three symbolic elements which have been used in the EPM's public communications, and became widely recognized by the public. The first symbolic element is the unique brand logo of the EPM, which consists a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element is the green color, which is strongly associated with taking care of the environment. The third element is the images of two high-profile celebrity comedians which were taken from advertising campaigns of the EPM. Further details about these symbolic elements are found in the supplementary appendix.

##*Sample selection and recruitment*

The survey was conducted in late January 2018. To capture natural variation in the personal relevance of the air-pollution policy, I sampled citizens from the major cities in the Haifa-Bay area (*Haifa*, *Nesher*, *Qiryat-Hayiim*, *Qiryat-Bialik*, *Qiryat-Mozkin*, *Qiryat-Ata* and *Qiryat-Yam*), and matched them with citizens from cities with a similar profile and population size at the center of the country (*Natania*, *Petach-Tikva* and *Rishon-Lezion*). I will henceforth refer to these two groups of cities as Haifa-bay and center.

I conducted the survey experiment through Qualtrics Survey Software. To select and recruit residents of these specific cities, I was assisted by an Israeli internet research panel company *iPanel*. The online survey link was sent by iPanel, between 28.1-1.2.2018. A total of `r N_0` respondents completed the survey (a response rate of `r round((N_0/5077)*100,1)`%).[^footnote_responserate] Of these, I later filtered out `r N_0-N_1` observations (`r round(((N_0-N_1)/N_0)*100,1)`%) due to multiple entries from the same IP address (n=`r filter_IP`), failure at the instructional manipulation test (n=`r filter_IMC`), submitting the survey in less than 3 or more than 30 minutes (n=`r filter_time`), and respondents under the age of 18 (n=`r filter_age`). These filters are not significantly associated with the areas or the experimental conditions. Ultimately, following this rigorous screening, I ended up with a sizable sample of `r N_1` subjects, of whom `r SVIVA2_01_haifa%>%nrow()` live in the Haifa-bay cities, and `r SVIVA2_01_center%>%nrow()` live in the center cities.

[^footnote_responserate]: According to iPanel's reporting, 5,077 people were invited to fill the survey. Of these people, approximately 330 women entered the survey, but were eventually denied of access, to avoid the over-representation of women in the sample. Hence, the response rate, when reducing those who were denied of access, is slightly higher  and reaches `r round((N_0)/(5077-330)*100,1)`%.

To disguise the purpose of the experiment, the survey was presented to the participants as a study about citizens' attitudes regarding environmental policies in general, as opposed to attitudes about EPM or the specific two policies. The participants were first asked pre-manipulation questions about their political ideology and their trust in government ministries in general and evaluation of their performance. Thereafter, they were asked about their interest in environmental policy issues, the extent to which they follow these issues in the media, their preference for more/less government investment in this field. They were also asked to describe the most salient environmental issue in their particular area of residence.[^footnote_relevance_manipulation]

[^footnote_relevance_manipulation]: The latter questions about environmental policy was initially designed as a manipulation of for perceived personal relevance (specifically for Haifa-Bay subjects with regard to the air-pollution policy). The questions which are mentioned above were included in the treatment group, whereas subjects in the control group were asked instead equivalent questions about their field of occupation. However, manipulation checks suggested that these priming questions did not enhance the perceived personal relevance of these policies for the participants (either from Haifa-Bay or center cities). Therefore, I will not regard to this treatment in the paper.

Thereafter, participants were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

The air-pollution and recycling policy plans were presented to each subject in a random order. Per each policy, the subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) and weak policy (i.e. logically unpersuasive). The policy activities which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles in the Haifa-Bay area. The strong recycling policy regarded increasing the supervision of manufacturers and reducing the use of disposable bags. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures which are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. Accordingly, the weak policy plans regarded reducing of regulatory burden on the factories/ manufacturers, and raising citizens' awareness of the efforts to improve air quality/reduce waste. The full texts of the weak and strong conditions for the two policy plans is displayed in Appendix A. For each subject, one of the two policies was strong and the other was weak, which enables me to compare both between- and within-subjects. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa.

The participants are also randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment, control no-symbols and control fake-symbols. In the symbol treatment condition, the two policies (again, one weak and one strong) are displayed with the above-mentioned three symbolic elements: the titles and the footer are colored with green, EPM logo is displayed at the top of the page, and the celebrities' images from the campaigns are attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags is attached to the recycling policy, given the relevance of the campaign to the policy plan; Friedman's image from the campaign of "starting to think green", which was more general and regarded to a variety of environmental issues, was attached to the air-pollution policy. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities images. Finally, in the control fake-symbols condition, the green color is replaced with blue, the EPM logo is replaced by a fake logo and the images of celebrities from the campaigns are replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition is to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols to that they will resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which can be attached to the government ministry and environmental issues. In that sense, it can be regarded as a placebo. In two prior surveys, I empirically confirmed that the fake symbols are unfamiliar, do not have strong positive or negative affect, and are not strongly associated with environment policy and EPM (supplementary appendix ##TBA##). The symbols conditions are displayed in the Appendix B.

Each policy plan was followed by questions about respondents' trust in the policy. Afterwards, the subjects were asked about their perceived level of elaboration on the last policy that was presented to them, conducted a memory test. They were also asked about the extent to which they perceive the two policies as personally relevant to themselves. Thereafter, I included an instructional manipulation check (IMC) item, which aims to assess participants’ diligence. After a set of demographic questions, I conducted a manipulation check, in which I asked the participants whether they recognize the Ministry's logo, and the two images of the celebrities. Finally, the participants are debriefed about the manipulations, and the possible deception. An English translation of the full survey
is available from the supplementary appendix.

Altogether, for each policy plan, the survey experiment has a factorial design of 3 (symbols: real symbols, fake-symbols, no-symbols) x 2 (information: strong vs. weak policy). The experimental groups are fairly balanced with respect to demographics, pre-manipulation variables and the two areas. The assignment to the different experimental groups is graphically presented in Figure 3.
